|

### 2023-07-11 paper/03-methods.Rmd
update fixed effects description #11 and add a brief description about resampling strategy for boostrapping uncertainty for random effects

In addition to the average STRAIN over all frames played, we can also calculate pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
\begin{equation}
\overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
\end{equation}
While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure].
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].

@@

>>>>

In addition to the average STRAIN over all frames played, we can also calculate pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
\begin{equation}
\overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
\end{equation}
While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved - both at the individual and team level.
Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure]. We also suspect that pass rusher's STRAIN is affected by their defensive teammates as well. 
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players and team involved, amongst other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + D_{d[i]} + O_{o[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of teams}, \\
O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of teams}.
\end{aligned}
\end{equation}

We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means, which is a useful property this problem's context since we do not observe the same number of plays for each player. For the team effects, this provides us with the average defense and offense team-level effects on a pass rusher's STRAIN. Due to the nested nature of players on teams, the individual pass rusher and blocker random intercepts reflect the respective player's effect relative to their team effects. We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

With regards to the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN. First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects. Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles. These are encoded as indicator variables with defensive ends and tackles as the reference groups for the pass rushers and blockers respectively. We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN. Finally, we also adjust for the length of the play (i.e., the elapsed time between the ball snap and a quarterback event). We expect a longer display to display lower STRAIN since this would indicate a quarterback has more time to throw the ball. NEED TO THINK ABOUT THIS MORE....

<!--Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].-->

@@

In order to provide a measure of uncertainty regarding about our random effects, we use a bootstrapping strategy similar to the approach in @yurko2019nflwar. Specifically, we resample team drives within games. This preserves the fact that team schedules are fixed, but allows for random variation in player usage since this is dependent on team decision making. By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays. For each resampled dataset, we fit the aforementioned multilevel model to generate a distribution of estimates for the considered player and team effects.
